---
title: "Intermediate Elective 2"
author: June McNally
date: 2026-05-10
format: pdf
execute:
  warning: false
  message: false
---


# Set Up

## Packages

```{r}
library(tidyverse)
library(readxl)
library(ggtext)
```
```{r}
library(tidyverse)
library(readxl)
library(ggtext)
library(here)
```


```{r}
aquatic_data <- read_excel("../data/aquatic_data.xlsx")
taxon_list <- read_csv("../data/taxon_list.csv")
```

#Problems:
I chose a South China Morning Post–style timeseries line chart, inspired by Steven Ponce's "Coal falls. Renewables rise." visualization, which tracks how proportions of different categories change over time on a dark background with colored lines.


The Aquatic Insects sheet spans 2020–2025 with counts of invertebrate taxa across multiple sites, making a proportion timeseries ideal for showing how the relative community composition shifts year over year.


Year is mapped to the x-axis, proportion of total count to the y-axis, and taxon group (Copepod, Ostracod, Cladocera, Amphipod, Chironomid) to line color.

```{r}
insects <- read_excel("../data/aquatic_data.xlsx",
                      sheet = "Aquatic Insects")
```

```{r}
aquatic_data <- read_excel(here("data", "aquatic_data.xlsx"))
taxon_list <- read_csv(here("data", "taxon_list.csv"))
```

```{r}
insects <- read_excel(here("data", "aquatic_data.xlsx"), 
                      sheet = "Aquatic Insects")
```

