---
title: "Intermediate Elective 2"
author: June McNally
date: 2026-05-10
format: pdf
execute:
  warning: false
  message: false
---


# Set Up

## Packages

```{r}
library(tidyverse)
library(readxl)
library(ggtext)
```
```{r}
library(tidyverse)
library(readxl)
library(ggtext)
library(here)
```


```{r}
aquatic_data <- read_excel("../data/aquatic_data.xlsx")
taxon_list <- read_csv("../data/taxon_list.csv")
```

#Problems:
1. Explain Your Inspriation
I chose a South China Morning Post–style timeseries line chart, inspired by Steven Ponce's "Coal falls. Renewables rise." visualization, which tracks how proportions of different categories change over time on a dark background with colored lines.


The Aquatic Insects sheet spans 2020–2025 with counts of invertebrate taxa across multiple sites, making a proportion timeseries ideal for showing how the relative community composition shifts year over year.


Year is mapped to the x-axis, proportion of total count to the y-axis, and taxon group (Copepod, Ostracod, Cladocera, Amphipod, Chironomid) to line color.

2. Plan Your Figure
## Sketch

![My figure sketIMG_6449.jpeg)

```{r}
insects <- read_excel("../data/aquatic_data.xlsx",
                      sheet = "Aquatic Insects")
```

```{r}
aquatic_data <- read_excel(here("data", "aquatic_data.xlsx"))
taxon_list <- read_csv(here("data", "taxon_list.csv"))
```

```{r}
insects <- read_excel(here("data", "aquatic_data.xlsx"), 
                      sheet = "Aquatic Insects")
```
```{r}
```{r}
# Read the Aquatic Insects sheet from the Excel file
insects <- read_excel(here("data", "aquatic_data.xlsx"),
                      sheet = "Aquatic Insects")

# Define the five key taxa we want to visualize
key_taxa <- c("Copepod", "Ostracod", "Cladocera", 
              "Amphipod", "Diptera: Chironomid ")

# Wrangle data: pivot to long format and calculate 
# proportion of each taxon per year
plot_data <- insects %>%
  select(`Date on Vial`, all_of(key_taxa)) %>%
  rename(Date = `Date on Vial`) %>%
  filter(!is.na(Date)) %>%
  mutate(Date = as.Date(Date), Year = year(Date)) %>%
  pivot_longer(cols = all_of(key_taxa), 
               names_to = "taxon", values_to = "count") %>%
  mutate(count = replace_na(count, 0),
         taxon = str_trim(taxon),
         taxon = if_else(taxon == "Diptera: Chironomid", 
                         "Chironomid", taxon)) %>%
  group_by(Year, taxon) %>%
  summarise(total = sum(count, na.rm = TRUE), .groups = "drop") %>%
  group_by(Year) %>%
  mutate(proportion = total / sum(total) * 100) %>%
  ungroup() %>%
  filter(Year >= 2020, Year <= 2025)

# Define colors for each taxon
colors <- c("Copepod" = "#E63946", "Ostracod" = "#2A9D8F",
            "Cladocera" = "#E9C46A", "Amphipod" = "#F4A261",
            "Chironomid" = "#457B9D")

# Get the last year's values for direct line labeling
labels <- plot_data %>% filter(Year == max(Year))

# Build the plot
ggplot(plot_data, aes(x = Year, y = proportion, color = taxon)) +
  # Add lines and points for each taxon
  geom_line(linewidth = 1.4) +
  geom_point(size = 2.5) +
  # Add direct labels at end of each line instead of a legend
  geom_text(data = labels, aes(label = taxon),
            hjust = -0.1, size = 3.2, fontface = "bold") +
  # Apply custom colors
  scale_color_manual(values = colors) +
  # Format x and y axes
  scale_x_continuous(breaks = 2020:2025,
                     expand = expansion(mult = c(0.02, 0.25))) +
  scale_y_continuous(labels = scales::label_number(suffix = "%")) +
  # Add titles and labels
  labs(title = "**Aquatic invertebrate composition over time**",
       subtitle = "Proportion of five key taxa, 2020–2025",
       x = NULL, y = "% of total count",
       caption = "Data: Goleta Slough Aquatic Sampling") +
  # Apply dark theme styling
  theme_minimal(base_size = 12) +
  theme(
    plot.background  = element_rect(fill = "#0D1B2A", color = NA),
    panel.background = element_rect(fill = "#0D1B2A", color = NA),
    text             = element_text(color = "white"),
    plot.title       = element_markdown(size = 14, color = "white"),
    plot.subtitle    = element_text(color = "#AAAAAA", size = 10),
    plot.caption     = element_text(color = "#666666", size = 8),
    axis.text        = element_text(color = "#AAAAAA"),
    panel.grid.major = element_line(color = "#1E3A5F"),
    panel.grid.minor = element_blank(),
    legend.position  = "none"
  )

# Save the final figure as a png file
ggsave(here("code", "aquatic_viz.png"), width = 8, height = 5, dpi = 300)
```
```

